---
title: "R Notebook"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Installing required R packages

```{r packages}
options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("rlang")
pkgs <- c("movMF", "Directional", "ggplot2", "dplyr", "tidyr", "rgl")
library(future.apply)
library(progressr)

plan(multisession, workers = 4)

handlers("txtprogressbar")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, require, character.only = TRUE))

has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (has_rgl) library(rgl)
```

# Setting seed

```{r seed}
set.seed(123)
```

# Basic helper functions

```{r Basic helper functions}
# Normalize rows of a matrix to unit length
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norms <- sqrt(rowSums(X^2))
  return(X/norms)
}

# Check unit length (returns summary)
check_unit_length <- function(X) {
  r <- sqrt(rowSums(as.matrix(X)^2))
  return(summary(r))
}

# Angular distance between two unit vectors (in radians and degrees)
angle_between <- function(u, v) {
  u <- as.numeric(u) / sqrt(sum(u^2))
  v <- as.numeric(v) / sqrt(sum(v^2))
  val <- acos(max(min(sum(u*v), 1), -1))
  return(c(radians = val, degrees = val * 180 / pi))
}

# Cosine similarity metric that measures how tightly points in a cluster stick to their mean direction
within_cosine <- function(X, cluster) {
  cluster <- as.integer(factor(cluster))
  K <- max(cluster)
  out <- numeric(K)
  
  for (k in 1:K) {
    idx <- which(cluster == k)
    if (length(idx) <= 1) next
    
    mu_hat <- colMeans(X[idx, , drop = FALSE])
    mu_hat <- mu_hat / sqrt(sum(mu_hat^2))
    out[k] <- mean(X[idx, , drop = FALSE] %*% mu_hat)
  }
  
  mean(out, na.rm = TRUE)
}

# Fast 2D projections for 3D spherical data
plot_projections <- function(X, label = NULL, title_prefix = "Projection") {
  X <- as.data.frame(X)
  colnames(X) <- paste0("x", 1:ncol(X))
  if (!is.null(label)) X$cluster <- factor(label)

  p1 <- ggplot(X, aes(x = x1, y = x2, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x2")) +
    theme_bw()

  p2 <- ggplot(X, aes(x = x1, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x3")) +
    theme_bw()

  p3 <- ggplot(X, aes(x = x2, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x2 vs x3")) +
    theme_bw()

  return(list(p12 = p1, p13 = p2, p23 = p3))
}

# Simple simulator for helper function demonstration
simulate_vmf_mixture <- function(d = 3, K_true, n, w, mu_list, k_list, force_min_one){
  
  mu_list <- lapply(mu_list, function(m)m / sqrt(sum(m^2))) # Normalize mean directions
  
  if (force_min_one && n >= K_true) { # Forces >= 1 point to be in each cluster. Note that we do NOT do this for actual tests. Maybe we should??? 
    z_init <- sample(1:K_true)
    z_rest <- sample(1:K_true, size = n - K_true, replace = TRUE, prob = w)
    z <- c(z_init, z_rest)
  } else {
    z <- sample(1:K_true, size = n, replace = TRUE, prob = w)
  }
  
  X <- matrix(NA, nrow = n, ncol = d)
  for (k in 1:K_true) {
    nk <- sum(z == k)
    if (nk > 0) {
      X[z == k, ] <- movMF::rmovMF(nk, theta = mu_list[[k]]*k_list[k]) 
    }
  }
  
  check_unit_length(X)
  table(z)
  
  X <- normalize_rows(X)
  list(X = X, z = z)
}
```

# Introduction

Directional data is concerned with unit vectors observations on a
circular plane or three dimensional space, rendering standard univariate
and multivariate analysis methods obsolete (Mardia, XIX). Directional
data is encountered in fields such as bioinformatics, collaborative
filtering (Banerjee, 1346), and the analysis of text documents
(Banerjee, 1348). Directional data is often extended into three
dimensions, with practical use of directional data arising in several
situations. The directions of magnetism in rocks, directions from the
Earth to stars, and directions of optical axes in quartz crystals are
all best modeled using spherical statistics (Mardia, 175). \n Of
particular importance to directional statistics is the von-Mises
distribution, which is perhaps one of the most useful distributions for
statistical inference the distribution most useful for statistical
inference (Mardia, 36). The vMF distribution is analogous to the
well-known Gaussian, or normal distribution (Banerjee, 1350 and Sra,
179), sharing many common properties. Similarly to the normal
distribution, the vMF distribution takes in two main parameters. one,
the mean direction μ, which is the average point at which vectors are
concentrated, and two, the concentration parameter, k, which is how
tightly points are clustering around that mean direction.

Despite being one of the most basic probability distributions for
multi-dimensional directional statistics, estimating the maximum
likelihood for the vMF distribution is difficult due to a transcendental
(non closed form) equation that needs to be solved in order to compute
the concentration parameter (Sra, 1). This parameter can be approximated
through several means, namely the Banerjee, Tanabe, and Newton methods.
These methods have various advantages and disadvantages. The Banerjee
method has the advantage of not using function evaluations, and
therefore being the fastest out of the three; however, it’s accuracy is
the least reliable. The Tanabe and Newton methods both are significantly
slower, but more accurate. In particular, the Newton method is the most
accurate.

Of particular importance to analyzing vMF models is mixing multiple vMF
distributions, which allows for multiple clusters of data that point in
different directions with varying levels of concentration. Clustering
algorithms often use the generative model framework, applying the
expectation maximization algorithm. Estimating the concentration
parameter k for each of these individual components has been of crucial
importance, as it would yield better results, particularly for difficult
clustering tasks in higher dimensional spaces (Banerjee, 1345).

In order to fit these vMF models (estimate the parameters of them in
order to best explain the data), the Expectation-Maximization, or EM
algorithm, is used. vMF models that only have a single distribution do
not require the EM algorithm; however, the EM algorithm becomes crucial
when there are clusters of data, meaning various vMF models are mixed
together. The expectation stage of the EM model computes the probability
each data point came from a certain cluster, then during the
maximization stage attempts to best update the parameters of each vMF
distribution in order to best fit the data (Hornik). In Hornik and
Grun’s movMF, an R package using the EM algorithm has been created with
the aim of fitting vMF mixtures: “The R package movMF contains
functionality to draw samples from finite mixtures of von Mises-Fisher
distributions and to fit these models using the expectation-maximization
algorithm for maximum likelihood estimation” (Hornik, 1). Various
examples of use are given, such as clustering text documents (Hornik,
24), which can group components within a document based on the detection
of certain keywords.

In this project, implementation of the EM algorithm will be done using
the movRF package. Various sets of vMF data will be analyzed and
clustered, with mean direction and k-values being estimated. Comparisons
to Euclidean k-value clustering techniques will be drawn, and clusters
will be visualized on the unit sphere for demonstration. Overall, we
will aim to analyze how well the movRF package implements the EM
algorithm for analyzing vMF data, and conduct several case studies using
various data sets.

## Simulation: Generating vMF Data in 3D

In order to a handle on what simulated vMF data is like, we create and
plot a basic data set that shows what a vMF distribution looks like.

The parameters we are working with are as follows:

**mu**: The mean direction unit vectors are pointing in. This cannot be
made more "difficult", and all mu values will be of equal complexity for
movMF to work with.

**kappa**: The measure of how strongly data is concentrated around the
mean direction. A lower value means lower concentration, making it more
difficult for movMF to estimate the mean direction. A value of 0 would
even mean complete dispersion of unit vectors, with no adherence to the
mean direction whatsoever. A higher concentration means increasing
adherence to the mean direction.

**n**: The number of unit vectors (points) we are working with. Higher
values of n means a greater sample size for movMF to work with, creating
more ease of estimating the mean direction.

# Creating basic parameter set

```{r basic params}
# A very optimal set of parameters for a vMF analysis. 
mu_base <- c(0, 0, 1) # mean direction
kappa_base <- 10 # concentration of 10 (pretty low)
n_base <- 200 # 200 total points

# Normalizing base parameters
mu_base_norm <- sqrt(sum(mu_base^2))
mu_base_norm

# Checking if mu is normalized. If not, it is normalized
mu_base <- mu_base / sqrt(sum(mu_base^2))
mu_base
```

# Simulating base vMF distribution data

This will simulate the hundred data points requested in the base vMF
data. We only print the head of data created, so six points are visible.
Printing the length of the data divided by three confirms 200 points
were created, with three coordinates each.

```{r Data Simulation: Base}
X_base <- rmovMF(n_base, theta = mu_base * kappa_base)
head(X_base)
print("length: " )
print(length(X_base)/3)
```

We need to always sanity check this because vectors must be unit length.
If not, k estimations will not function.

```{r Sanity Check: X_base Unit Length}
lens_base <- sqrt(rowSums(X_base^2))
summary(lens_base)
```

## Visualization

We can get a good graph on what this data looks like by visualizing it,
both on the three 2D planes and a 3D representation of the unit sphere.

# 2D planes

```{r plot-xy: base}
df <- data.frame(x = X_base[,1], y = X_base[,2], z = X_base[,3])

ggplot(df, aes(x, y)) +
  geom_point(alpha = 0.6) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Projection of base vMF data Samples (x vs y)", x = "x", y = "y")
```

```{r plot-xz: base}
df <- data.frame(x = X_base[,1], y = X_base[,2], z = X_base[,3])

ggplot(df, aes(x, z)) +
  geom_point(alpha = 0.6) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Projection of base vMF data Samples (x vs z)", x = "x")
```

```{r plot-yz: base}
df <- data.frame(x = X_base[,1], y = X_base[,2], z = X_base[,3])

ggplot(df, aes(y, z)) +
  geom_point(alpha = 0.6) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Projection of base vMF data Samples (y vs z)", y = "y")
```

# 3D projection

```{r base plot-3d}
library(rgl)
open3d()
plot3d(X_base, col = "blue", size = 5)
rglwidget()
```

## Parameter estimation

With parameter estimation, we can attempt to estimate mean and
concentration without knowing beforehand.

```{r base parameter estimation}
# Estimating base mean
mu_base_hat_raw <- colMeans(X_base)
mu_base_hat <- mu_base_hat_raw / sqrt(sum(mu_base_hat_raw^2))
mu_base_hat
```

Compare the estimated mean [0.044, 0.059, 0.997] to the true mean [0, 0,
1].

The exact degree of error between true mu and the estimated mu can be
measured by taking angle between vectors.

```{r base mu-comparison}
mu_base_true <- mu_base

# Angle between vectors
angle_base_rad <- acos(sum(mu_base_true * mu_base_hat))
angle_base_deg <- angle_base_rad * 180 / pi

list(
  mu_base_true = mu_base_true,
  mu_base_hat = mu_base_hat,
  base_angle_degrees= angle_base_deg
)
```

# Concentration estimation

movMF can attempt to estimate the concentration variable K, which is the
foremost variable being studied and approximated.

```{r base k estimation}
# Setting k to 1 fits a single vMF distribution. Therefore, the E step in the EM algorithm is skipped - every point will belong to the same component. Only the M step is used.
fit1 <- movMF(X_base, k = 1) # one mixture
fit1

kappa_base_hat <- norm(fit1$theta, "2")
mu_base_hat_mle <- fit1$theta/kappa_base_hat

list(
  # Estimation of mean
  mu_base_hat_mle = mu_base_hat_mle,
  # Estimation of kappa
  kappa_base_hat = kappa_base_hat
)
```

The question remains of what tend to throw off the algorithm more - is
it sample size, or the concentration? These questions can be further
investigated by the reader in the Phase 2 file, which includes this
content and a further study into what effects the accuracy of movRF. For
the sake of brevity, this file will only go in depth on later studies
involving multiple mixtures.

## Multiple Mixtures

In this section, mixtures of vMF distributions are fit, rather than just
a single one. This is far more realistic for the work being done with
movMF, as seldom are we dealing with just a single mixture.

# Creating a basic mixture model

This example mixture model, five mixtures are created with equal
weights, and they are all very far away from each other. This is similar
to the "base" model we used in the single distrubtion model, where the
purpose of it is demonstration rather than rigorous testing.

```{r simulated mixture}
d <- 3 # Number of dimensions
K_true <- 5
n <- 500

# Mixture weights
w <- c(0.20, 0.20, 0.20, 0.20, 0.20)

# Component mean directions
mu_list <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1),
  c(1, 1, 0),
  c(1, 0, 1)
)
mu_list <- lapply(mu_list, function(m)m / sqrt(sum(m^2)))

# Component concentrations (kappas)
# A larger value means tighter clustering
k_list <- c(25, 40, 30, 15, 20)

# Sample component labels
z <- sample(1:K_true, size = n, replace = TRUE, prob = w)

# Generate data (rmovMF parameter: theta = mu*kappa)
X <- matrix(NA, nrow = n, ncol = d)
for (k in 1:K_true) {
  nk <- sum(z == k)
  if (nk > 0) {
    X[z == k, ] <- movMF::rmovMF(nk, theta = mu_list[[k]]*k_list[k]) 
  }
}

# Sanity checks
check_unit_length(X)
table(z)
```

# Visualizing the mixture

# 2D planes

```{r phase3 projections}
plots <- plot_projections(X, label = z, title_prefix = "True mixture (simulated)")
plots$p12
plots$p13
plots$p23
```

# 3D projection

```{r phase3-3d, eval=TRUE}
# Interactive 3D plot
rgl::open3d()
cols <- as.integer(factor(z))
rgl::plot3d(X, col = cols, size = 5, type = "s")
rgl::title3d("Simulated mixture on the unit sphere")
rgl::rglwidget()
```

## Attempting to fix mixtures of vMF for multiple mixtures, and selecting K (estimated number of mixtures)

We will fit vMF mixtures, and compute BIC (grid over K). BIC is used for
model selection, and helps us determine how well a model explains the
data.

```{r fit-grid}
K_grid <- 1:8
fits <- vector("list", length(K_grid))
names(fits) <- paste0("K", K_grid)

# Data must be in a matrix. 

X_unit <- normalize_rows(X)

for (K in K_grid) {
  fits[[paste0("K", K)]] <- movMF::movMF(X_unit, k = K, nruns = 10)
}

# Extract, store BIC values
bic_df <- data.frame(
  K = K_grid,
  BIC = sapply(fits, BIC)
)

bic_df
```

BIC is determining what amount of models the data most likely has,with a
lower BIC being better. In this case, BIC shows that 5 is the best fit.

This chosen K makes sense, as we did indeed select 5 mixtures to work
with earlier. However, the BIC was nearly flat, which indicate that
movRF was starting to having some trouble.

```{r choose-K}
K_hat <- bic_df$K[which.min(bic_df$BIC)]
K_hat
best_fit <- fits[[paste0("K", K_hat)]]
```

```{r bic-plot}
ggplot(bic_df, aes(x = K, y = BIC)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(title = "Model selection for vMF mixtures", subtitle = "Choose K with the smallest BIC", x = "Number of components (K)", y = "BIC")
```

We will now estimate this mixture with K-means and movMF, and see how
much the two models agree or disagree.

## Estimating mixture model with K-means

```{r kmeans-euclid}
km <- kmeans(X_unit, centers = K_hat, nstart = 25)
km_cluster <- factor(km$cluster)

table(km_cluster)
```

```{r km-projections}
plots_km <- plot_projections(X_unit, label = km_cluster, title_prefix = paste0("Euclidean k-means (K = ", K_hat, ")"))
plots_km$p12
plots_km$p13
plots_km$p23
```

## Estimating with movMF

```{r vmf-clusters}
vmf_cluster <- predict(best_fit)
vmf_cluster <- factor(vmf_cluster)
table(vmf_cluster)
```

```{r vmf-projections}
plots_vmf <- plot_projections(X_unit, label = vmf_cluster, title_prefix = paste0("vMF mixture clustering (K = ", K_hat, ")"))
plots_vmf$p12
plots_vmf$p13
plots_vmf$p23
```

Upon close inspection, some of the dots in the border areas between
clusters do indeed switch from cluster to cluster depending on the
algorithm used. An agreement matrix shows how many dots the algorithms
put into a cluster itself:

```{r agreement}
# Simple agreement table (labels might be permuted, so interpret carefully)
table(vmf_cluster, km_cluster)
```

Although there is some disagreement, for the most part we see that four
clear clusters are established. However, there is some demonstrable
difference between the estimates of both algorithms; which one is doing
better?

# Final Study

In this analysis, a large scale comparison between the effectiveness of
Euclidian (K-means) and Spherical Analysis (movMF) models is conducted
on multi-mixture models. This is done via comparing against several
variables:

1.  Total number of mixtures: The number of mixtures being fitted in
    each dataset.

2.  Sample size: The total number of points in each dataset.

3.  Average concentration: The average concentration of mixtures.

Three more variables represent the variance of mixtures, concentrations,
and means:

1.  Variance of means: Means always being exactly 90 degrees apart from
    each other wouldn't be too realistic. In reality, most mixture
    models will be far less optimal, and some clusters are likely to
    overlap to varying degrees. In every one of these helper functions
    range from 1 (most optimal) to 4 (difficult for estimation and very
    unoptimal)

```{r mean variance helper function}
# Generates K mean directions, with separation of the directions and number of mean directions specified.
generate_mus <- function(K, d = 3, separation = 1, max_iter = 500) {
  
  target <- switch(
    as.character(separation),
    "1" = c(0.2, 0.4), # Vectors are very separated
    "2" = c(0.5, 0.7), # Vectors are less separated
    "3" = c(0.75, 0.9), # Vectors start to become more clustered
    "4" = c(0.9, 0.95) # Extremely clustered vectors
  )
  
  for (iter in 1:max_iter) {
    M <- matrix(rnorm(K * d), K, d) # Random matrix, K (amount of directions) * 3 dimensions
    M <- M / sqrt(rowSums(M^2)) # Normalize each row
    
    cos_vals <- (M %*% t(M))[upper.tri(M %*% t(M))] # Computes cosine-similarity matrix
    
    if (mean(cos_vals) >= target[1] && mean(cos_vals) <= target[2]) { # Checks if average of pairwise cosines lies inside our target interval. Only returns if it does.
      return(split(M, row(M)))
    }
  }
  
  split(M, row(M))
}
```

```{r variance of means helper function example}
K_true       <- 3
n            <- 500
fixed_w      <- c(0.33, 0.33, 0.34)   # almost equal weights
fixed_kappas <- rep(30, K_true)                       # equal concentrations

sims <- lapply(1:4, function(sep_level) {
  
  mu_list <- generate_mus(K = K_true, d = 3, separation = sep_level)
  
  sim <- simulate_vmf_mixture(
    d          = 3,
    K_true     = K_true,
    n          = n,
    w          = fixed_w,
    mu_list    = mu_list,
    k_list     = fixed_kappas,
    force_min_one = TRUE
  )
  
  list(X = sim$X, z = sim$z, sep_level = sep_level)
})

labels <- c(
  "1 = Very separated mean directions",
  "2 = Moderately separated",
  "3 = Starting to cluster",
  "4 = Extremely clustered (almost overlapping)"
)

# 2D projections
for (i in seq_along(sims)) {
  plots <- plot_projections(sims[[i]]$X,
  label = sims[[i]]$z,
  title_prefix = labels[i])
  print(plots$p12)
}

# 3D projections
for (i in seq_along(sims)) {
  rgl::open3d(useNULL = TRUE)
  cols <- as.integer(factor(sims[[i]]$z))
  rgl::plot3d(sims[[i]]$X, col = cols, size = 3, type = "s")
  rgl::title3d(main = labels[i])
  widget <- rgl::rglwidget()
  rgl::close3d()
  print(widget)
}
```

2.  Variance of concentrations: Concentrations should not all be the
    "average concentration"; this function controls how much the
    concentrations vary around the mean concentration. A dedicated
    helper function is used to generate these kappas.

```{r concentration variance helper function and example}
make_kappas <- function(K, kappa_mean, variation = 1) {
  
  # map variation level to log-scale spread
  sigma <- switch(
    as.character(variation),
    "1" = 0.2, # Creates a very tight cluster around mean
    "2" = 0.5, # Moderate spread around given mean
    "3" = 0.8, # Creates wide spread around given mean
    "4" = 1.1, # Very wide spread around mean
    stop("variation must be 1, 2, 3, or 4")
  )
  
  # log-mean so that expected value ≈ kappa_mean
  mu <- log(kappa_mean) - (sigma^2 / 2) # Guarantees positivity
  
  # sample kappas
  kappas <- rlnorm(K, meanlog = mu, sdlog = sigma)
  
  return(kappas)
}
```

```{r variance of concentrations example}
# Example of all four variances
K_true <- 3
n <- 500
w <- c(0.33, 0.33, 0.34) # Weights are uniform
mu_list <- list(c(1, 0, 0), c(0, 1, 0), c(0, 0, 1)) # Means are all equally distant from each other
kappa_mean <- 30

sims <- lapply(1:4, function(var_level){
  
  kappas <- make_kappas(K = K_true, 
                        kappa_mean = kappa_mean, 
                        variation = var_level)
  
  sim <- simulate_vmf_mixture(
    d       = 3,
    K_true  = K_true,
    n       = n,
    w       = fixed_w,
    mu_list = mu_list,
    k_list  = kappas,
    force_min_one = TRUE
  )
  list(X = sim$X, z = sim$z, var_level = var_level)
})

labels <- c(
  "1 = Very equal concentrations (tight cluster around mean)",
  "2 = Moderate concentration variation",
  "3 = High concentration variation",
  "4 = Very unequal concentrations (some very peaked, some very broad)"
)

# 2D projections
for (i in seq_along(sims)) {
  plots <- plot_projections(sims[[i]]$X,
  label = sims[[i]]$z,
  title_prefix = labels[i])
  print(plots$p12)
}

# 3D projections
for (i in seq_along(sims)) {
  rgl::open3d(useNULL = TRUE)
  cols <- as.integer(factor(sims[[i]]$z))
  rgl::plot3d(sims[[i]]$X, col = cols, size = 3, type = "s")
  rgl::title3d(labels[i])
  widget <- rgl::rglwidget()
  rgl::close3d()
  print(widget)
}
```

3.  Variance of sample sizes: Mixture sizes vary from all being the same
    size (portion of total sample size allocated to them), to highly
    varying in what portion of the total sample size is allocated to
    them.

```{r sample size variance example}
make_weights <- function(K, variation = 1) {
  
  # Variation of weights, samples from a Dirichlet distribution. 
  alpha <- switch(
    as.character(variation),
    "1" = 10, # Weights are very similar, almost uniform
    "2" = 2, 
    "3" = 0.7,
    "4" = 0.3, # Weights are very unequal, some are very large, others are near zero.
    stop("variation must be 1, 2, 3, or 4")
  )
  
  # Dirichlet via Gamma
  x <- rgamma(K, shape = alpha, rate = 1) # Draws K independent numbers from a gamma distribution
  w <- x / sum(x) # Normalizes each vector
  
  return(w)
}

# Example of all four variances
w <- c(0.33, 0.33, 0.34) # Weights are uniform
mu_list <- list(c(1, 0, 0), c(0, 1, 0), c(0, 0, 1)) # Means are all equally distant from each other
sims <- lapply(1:4, function(var_level){
  sim <- simulate_vmf_mixture(
    d = 3,
    K_true = 3,
    n = 500,
    w = make_weights(3, var_level),
    mu_list = mu_list,
    k_list = fixed_kappas,
    force_min_one = TRUE
  )
  list(X = sim$X, z = sim$z, var_level = var_level)
})

labels <- c("Very equal weights (almost uniform)",
            "Moderate weight variation",
            "High weight variation",
            "Very unequal weights")

for (i in seq_along(sims)) {
  plots <- plot_projections(sims[[i]]$X,
  label = sims[[i]]$z,
  title_prefix = labels[i])
  print(plots$p12)
}

for (i in seq_along(sims)) {
  rgl::open3d(useNULL = TRUE)
  cols <- as.integer(factor(sims[[i]]$z))
  rgl::plot3d(sims[[i]]$X, col = cols, size = 3, type = "s")
  rgl::title3d(labels[i])
  widget <- rgl::rglwidget()
  rgl::close3d()
  print(widget)
}

```

# Superfunction used for each mixture setting

```{r superfunction}
# This grid creates a massive simulation grid, and creates every possible combination of our variables.
# n: sample size. goes from 100, 300, ..., 1000.
# k: mean kappa. goes from 10, 30, ..., 100.
# three variance variables mentioned in helper functions: goes through each of the four levels.

simulate_complete <- function(number_mixtures) {
  grid <- expand.grid(
  n = seq(100, 1000, by = 200),
  kappa_mean = seq(10, 100, by = 20),
  weight_var = 1:4,
  kappa_var = 1:4,
  mu_sep = 1:4,
  rep = 1:25 # test is ran 25 times for a monte carlo experiment
) %>%
  as_tibble()

K <- number_mixtures # Number of mixtures (uses the only parameter, number_mixtures).
d <- 3 # Number of dimensions (always three)

with_progress({
  p <- progressor(steps = nrow(grid))
  
  results <- future_lapply(seq_len(nrow(grid)), function(i) {
    
    p(sprintf("Sim %d", i))
    row <- grid[i, ]
    
    # randomly generating the variables for our analysis, based on given weight variance, mean of kappas, and separation of mixtures.
    
    w <- make_weights(K, row$weight_var)
    kappas <- make_kappas(K, row$kappa_mean, row$kappa_var)
    mus <- generate_mus(K, d, row$mu_sep)
    
    # putting together entire dataset
   sim <- simulate_vmf_mixture(
        d       = d,
        K_true  = K,
        n       = row$n,
        w       = w,
        mu_list = mus,
        k_list  = kappas,
        force_min_one = TRUE
      )
    
   # using vMF to fit model
    vmf_fit <- tryCatch(
      movMF::movMF(sim$X, k = K, nruns = 3),
      error = function(e) NULL
    )
    
    vmf_cluster <- if (is.null(vmf_fit)) NA else predict(vmf_fit)
    
    vmf_score <- if (all(is.na(vmf_cluster))) NA else
      within_cosine(sim$X, vmf_cluster)
    
    # using old fashioned k-means to fit model
    km <- kmeans(sim$X, centers = K, nstart = 5)
    km_score <- within_cosine(sim$X, km$cluster)
    
    # stores all of our parameters, and score from two algorithms, in a dataframe
    data.frame(
      n = row$n,
      kappa_mean = row$kappa_mean,
      weight_var = row$weight_var,
      kappa_var = row$kappa_var,
      mu_sep = row$mu_sep,
      rep = row$rep,
      vmf_score = vmf_score,
      km_score = km_score
    )
    
  }, future.seed = TRUE)
  
  results <- bind_rows(results)
})

# compute how often vMF beat K-means
results <- results %>%
  mutate(
    diff = vmf_score - km_score,
    vmf_win = vmf_score > km_score
  )

# compute average score difference for each combination of variables
win_df <- results %>%
  group_by(n, kappa_mean, weight_var, kappa_var, mu_sep) %>%
  summarise(
    win_rate = mean(vmf_win, na.rm = TRUE),
    diff_mean = mean(diff, na.rm = TRUE),
    .groups = "drop"
  )

# then, for each level of μ separation, we plot it all into a winrate heatmap so we can see where vMF is doing better.

for (lvl in 1:4) { # go through each μ separation level
  
  plot_df <- win_df %>%
    filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0.5,
      name = "vMF win rate"
    ) +
    
    facet_grid(weight_var ~ kappa_var) +
    
    labs(
      title = "vMF vs k-means Performance",
      subtitle = paste("μ separation level =", lvl,
                       "| Rows: weight variation | Columns: kappa variation"),
      x = "Mean concentration (kappa)",
      y = "Sample size (n)"
    ) +
    
    theme_bw()
  
  print(p)
  }
  invisible(list(raw = results, summary = win_df))
}
```

# Two Mixtures

```{r double mixture simulation, cache=TRUE}
# We do NOT want to run this simulation every time the page is ran, or especially when it is knitted.
# Therefore we check if the simulation has been ran first. If it has, the simulation has been saved to a file.
# knittr then retrieves it and uses it, instead of running the simulation itself and likely crashing.
if (!file.exists("simulation_results_K2.rds")) {
  message("No saved results found — running full simulation...")
  result <- simulate_complete(2)
  saveRDS(result, "simulation_results_K2.rds")
} else {
  message("loading cached simulation results...")
  result <- readRDS("simulation_results_K2.rds")
  
  win_df <- result$summary

for (lvl in 1:4) {
  plot_df <- win_df %>% filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0.5, name = "vMF win rate") +
    facet_grid(weight_var ~ kappa_var) +
    labs(title = "vMF vs k-means Performance",
         subtitle = paste("μ separation level =", lvl,
                          "| Rows: weight variation | Columns: kappa variation"),
         x = "Mean concentration (kappa)",
         y = "Sample size (n)") +
    theme_bw()
  
  print(p)
  }
}
```

**Two mixtures analysis**

μ separation level 1 (more mixture separation): vMF has the edge with
very unconcentrated mixtures, particularly when the variance
concentration between the two mixtures is low as well. Also this is
perhaps a niche case, this is a good sign, as estimating mixtures is at
it's most difficult here.

μ separation level 2 (less mixture separation): Similar to level 1.
However, vMF isn't as dominant at very low sample sizes, but also gains
some traction with higher sample sizes. Beyond low kappa variation,
kmeans wins or the results are inconclusive.

μ separation level 3 (slight mixture clustering): A noted change from
the previous two μ separation levels. vMF does somewhat well with low
kappa variance when the mean concentration is 30-50, for all weight
variations besides high.

μ separation level 4 (high mixture clustering): Very similar to μ
separation level 3, albeit with the mean concentrations where vMF is
effective shifting right.

Overall Analysis: There seem to be two trends here. In the first two
lower μ clustering levels, vMF does well with models where the mixtures
have uniformly low concentration, and weights are not extremely varied.
In the next two higher μ separation levels, vMF remains effective with
low concentration variation, but more so with higher mean
concentrations.

# Three mixtures

```{r triple mixture simulation, cache=TRUE}
if (!file.exists("simulation_results_K3.rds")) {
  message("No saved results found — running full simulation...")
  result <- simulate_complete(3)
  saveRDS(result, "simulation_results_K3.rds")
} else {
  message("loading cached simulation results...")
  result <- readRDS("simulation_results_K3.rds")
  
  win_df <- result$summary

for (lvl in 1:4) {
  plot_df <- win_df %>% filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0.5, name = "vMF win rate") +
    facet_grid(weight_var ~ kappa_var) +
    labs(title = "vMF vs k-means Performance",
         subtitle = paste("μ separation level =", lvl,
                          "| Rows: weight variation | Columns: kappa variation"),
         x = "Mean concentration (kappa)",
         y = "Sample size (n)") +
    theme_bw()
  
  print(p)
  }
}
```

**Three mixtures analysis**

μ separation level 1 (more mixture separation): vMF has the edge with
very unconcentrated mixtures, particularly when the mean concentration
is low as well. Also this is perhaps a niche case, this is a good sign,
as estimating mixtures is at it's most difficult here.

μ separation level 2 (less mixture separation): Similar to level 1.
However, vMF only has a decisive edge with medium concentrations that
don't vary much from the mean. With high weight variation, this
advantage ceases.

μ separation level 3 (slight mixture clustering): vMF is does quite well
when concentrations don't vary much and concentration isn't low. This
trend continues until mixture weights highly vary; however, even in the
most varied sample, vMF managed to hold on with very low sample sizes.
This combination (low sample size, high weight variantion, AND low kappa
variation) is quite difficult, so even though it is a sliver of the
data, this shows a pain point where vMF is strong.

μ separation level 4 (high mixture clustering): vMF has mostly
completely lost the edge. With small sample sizes, sometimes it edges
out k-means, but it is uncommon and only occurs with high kappa
variation.

Overall Analysis: With three mixtures, vMF wins when concentrations are
not varied much. As the means of mixtures continue to cluster and get
closer to each other, vMF requires higher concentrations to keep
winning, until with high clustering it rarely has an advantage.

# Five mixtures

```{r five mixtures simuation, cache=TRUE}
if (!file.exists("simulation_results_K5.rds")) {
  message("No saved results found — running full simulation...")
  result <- simulate_complete(5)
  saveRDS(result, "simulation_results_K5.rds")
} else {
  message("loading cached simulation results...")
  result <- readRDS("simulation_results_K5.rds")
  
  win_df <- result$summary

for (lvl in 1:4) {
  plot_df <- win_df %>% filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0.5, name = "vMF win rate") +
    facet_grid(weight_var ~ kappa_var) +
    labs(title = "vMF vs k-means Performance",
         subtitle = paste("μ separation level =", lvl,
                          "| Rows: weight variation | Columns: kappa variation"),
         x = "Mean concentration (kappa)",
         y = "Sample size (n)") +
    theme_bw()
  
  print(p)
  }
}
```

**Five mixtures analysis**

μ separation level 1 (more mixture separation): vMF seems to have the
edge when concentrations aren't varied and are nearly uniform, and
weights are nearly uniform to moderately varied. In increasing
variations of concentrations, the efficacy of vMF vs K-means becomes
more contested.

μ separation level 2 (less mixture separation): Much like separation
level 1, but vMF is less dominant overall. Interestingly, K-means
completely dominates where mean concentration is very low. vMF has a
better shot at low sample size models.

μ separation level 3 (slight mixture clustering): Similiar to level 2.
vMF only has the edge with some smaller sample sizes.

μ separation level 4 (high mixture clustering): vMF has totally lost
most of it's edge, but seems to make a small resurgence with very low
concentrations when concentrations aren't varied.

Overall Analysis: vMF wins with edge cases, namely less clustering of
mixtures and lower weight and concentration variation.

# Seven mixtures

```{r seven mixtures simuation, cache=TRUE}
if (!file.exists("simulation_results_K7.rds")) {
  message("No saved results found — running full simulation...")
  result <- simulate_complete(7)
  saveRDS(result, "simulation_results_K7.rds")
} else {
  message("loading cached simulation results...")
  result <- readRDS("simulation_results_K7.rds")
  
  win_df <- result$summary

for (lvl in 1:4) {
  plot_df <- win_df %>% filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0.5, name = "vMF win rate") +
    facet_grid(weight_var ~ kappa_var) +
    labs(title = "vMF vs k-means Performance",
         subtitle = paste("μ separation level =", lvl,
                          "| Rows: weight variation | Columns: kappa variation"),
         x = "Mean concentration (kappa)",
         y = "Sample size (n)") +
    theme_bw()
  
  print(p)
  }
}
```

**Seven mixtures analysis**

μ separation level 1 (more mixture separation): vMF wins with low
concentration variation and higher mean concentrations. Interestingly,
it really exceeds with low concentration variation, low sample sizes,
and high weight variation.

μ separation level 2, 3, and 4: Trends become harder to note here. It
seems vMF has a good shot with low to medium concentration variation and
low to medium weight variation.

Overall Analysis: vmF sporadically wins low to medium concentration
variation and low to medium weight variation. It is hard to "prove" here
that vMF is the better option, and more testing may be needed.

# Conclusion

In all four mixture models studied, vMF tends to lose steam as
concentrations vary more from the mean concentration given. A recurring
trend, namely with two, three, and five mixture models is that vMF
excels with low concentration variance, but as mixtures cluster together
more and become less optimal, vMF shifts from doing well with low
concentrations to doing well with higher concentrations. In most tiles
where vMF vs K-means is inconclusive, K-means does better with higher
sample sizes, but as sample sizes decrease, vMF occasionally gains the
edge, particularly with very low sample sizes.

# Limitations and future work

Obviously I would've liked to have gone more in depth with this,
exploring the limits of more variables, and testing extreme edge cases
where both models are going to struggle and the strengths of one model
are more likely to make it come out on top. However, due to time and
computing constraints, a general experiment across at a wide range of
mixture models was done. Also, more experiments would've assured that
this experiment was more accurate and less prone to outliers skewing the
data; 70 trials compared to 25 would give more peace of mind.

Further directions this study could go in include:

-   A great next stage could be some sort of resulting function that
    takes in nearly any mixture model, and predicts whether vMF or
    K-means is the better choice.

-   Nearly every practical range of variables tested, some delineation
    of when accurate testing is impossible established.

-   Perhaps a "smoother" heatmap that doesn't even have tiling, but a
    smooth transition between colors that indicate whether vMF or
    K-means is winning. However, this would involved far more testing to
    be smooth, as well as far more repetitions to accurately tie down
    scores.

-   Perhaps testing in more dimensions, if applicable. Regardless,
    visualization could become far more difficult at this stage, and a
    function predicting whether vMF or K-means is more applicable might
    be the only helpful outcome.

# References

Mardia, K. V. and Jupp, P. E. (2000). Directional Statistics. Wiley.

Banerjee, A., Dhillon, I. S., Ghosh, J., and Sra, S. (2005). Clustering
on the Unit Hypersphere using von Mises–Fisher Distributions.

Sra, S. (2012). A short note on parameter approximation for von
Mises–Fisher distributions.

Hornik, K. and Grün, B. (2014). movMF: An R package for fitting mixtures
of von Mises–Fisher distributions.
