---
title: "R Notebook"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this analysis, a large scale comparison between the effectiveness of Euclidian and Spherical Analysis models is conducted. This is done via comparing against several variables:

1. Total number of mixtures: The number of mixtures being fitted in each dataset. 
2. Sample size: The total number of points in each dataset. 
3. Average concentration: The average concentration of mixtures.

Three more variables represent the variance of mixtures, concentrations, and means:
1. Variance of means: Means always being 90 degrees apart from each other wouldn't be too realistic. 
2. Variance of concentrations: Concentrations should not all be the "average concentration"; this controls how much the concentrations vary around the mean concentration
3. Variance of sample sizes: Mixture sizes vary from all being the same size (portion of total sample size allocated to them), to highly varying in what portion of the total sample size is allocated to them.

In order to analyze these how variables, 

```{r helper functions}

### generate_mus
# This function generates the vector means for mixtures in datasets.
# A "perfect" dataset would have mixtures centered far away from each other; the means would look like (0, 0, 1), (1, 0, -1) etc. Obviously, this isn't realistic, and we have to consider that vector means might be close to each other, and that mixtures might nearly be overlapping.
# This function aims to simulate the challenges of this parameter, and creates sets of means according to cosine similarity.
# Cosine similiarity is simply the cosine of angles between two vectors
generate_mus <- function(K, d = 3, separation = 1, max_iter = 1000) {
  
  # Upper bounds on cosine similarity (i.e., enforce minimum angle)
  max_cos_allowed <- switch(
    as.character(separation),
    "1" = 0.4,   # well separated
    "2" = 0.7,   # moderate
    "3" = 0.9    # weak separation
  )
  
  for (iter in 1:max_iter) {
    M <- matrix(rnorm(K * d), K, d)
    M <- M / sqrt(rowSums(M^2))
    
    cos_mat <- M %*% t(M)
    cos_vals <- cos_mat[upper.tri(cos_mat)]
    
    # Enforce minimum separation via max cosine similarity
    if (max(cos_vals) <= max_cos_allowed) {
      return(split(M, row(M)))
    }
  }
  
  warning("Did not meet separation constraint; returning last sample.")
  split(M, row(M))
}
generate_mus(3, 3, 1)

generate_mus(3, 3, 2)

generate_mus(3, 3, 3)
```