---
title: "R Notebook"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this file, we will go through stage 3 and 4 of the vMF clustering
project.

-   **Phase 3: Mixture Modeling and Clustering** Mixtures of vMF
    distributions will be fit, rather than a single distribution like in
    stage 2. The number of clusters will be selected via BIC,
    visualization on the unit sphere will be conducted, and comparisons
    with Euclidian K-means will be done.
-   **Phase 4: Comparative Study** Simulations of Euclidian, traditional
    analysis models vs spherical models will be conducted.

# Installing required packages

```{r packages}
pkgs <- c("movMF", "Directional", "ggplot2", "dplyr", "tidyr", "rgl", "future.apply", "progressr")
install.packages(c("future.apply", "progressr"))

library(future.apply)
library(progressr)
library(dplyr)

to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, require, character.only = TRUE))

has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (has_rgl) library(rgl)
```

```{r seed}
set.seed(123)
```

# Helper functions for phase 3 and 4

```{r helpers}
# Normalize rows of a matrix to unit length
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norms <- sqrt(rowSums(X^2))
  return(X/norms)
}

# Check unit length (returns summary)
check_unit_length <- function(X) {
  r <- sqrt(rowSums(as.matrix(X)^2))
  return(summary(r))
}

# Angular distance between two unit vectors (in radians and degrees)
angle_between <- function(u, v) {
  u <- as.numeric(u) / sqrt(sum(u^2))
  v <- as.numeric(v) / sqrt(sum(v^2))
  val <- acos(max(min(sum(u*v), 1), -1))
  return(c(radians = val, degrees = val * 180 / pi))
}

# Fast 2D projections for 3D spherical data
plot_projections <- function(X, label = NULL, title_prefix = "Projection") {
  X <- as.data.frame(X)
  colnames(X) <- paste0("x", 1:ncol(X))
  if (!is.null(label)) X$cluster <- factor(label)

  p1 <- ggplot(X, aes(x = x1, y = x2, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x2")) +
    theme_bw()

  p2 <- ggplot(X, aes(x = x1, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x3")) +
    theme_bw()

  p3 <- ggplot(X, aes(x = x2, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x2 vs x3")) +
    theme_bw()

  return(list(p12 = p1, p13 = p2, p23 = p3))
}
```

------------------------------------------------------------------------

# Phase 3: Mixture Models and Clustering

We will create a simulated mixture of vMF distributions:

```{r simulated mixture}
d <- 3 # Number of dimensions
K_true <- 5
n <- 500

# Mixture weights
w <- c(0.20, 0.20, 0.20, 0.20, 0.20)

# Component mean directions
mu_list <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1),
  c(1, 1, 0),
  c(1, 0, 1)
)
mu_list <- lapply(mu_list, function(m)m / sqrt(sum(m^2)))

# Component concentrations (kappas)
# A larger value means tighter clustering
k_list <- c(25, 40, 30, 15, 20)

# Sample component labels
z <- sample(1:K_true, size = n, replace = TRUE, prob = w)

# Generate data (rmovMF parameter: theta = mu*kappa)
X <- matrix(NA, nrow = n, ncol = d)
for (k in 1:K_true) {
  nk <- sum(z == k)
  if (nk > 0) {
    X[z == k, ] <- movMF::rmovMF(nk, theta = mu_list[[k]]*k_list[k]) 
  }
}

# Sanity checks
check_unit_length(X)
table(z)
```

# Visualizing the mixture

```{r phase3 projections}
plots <- plot_projections(X, label = z, title_prefix = "True mixture (simulated)")
plots$p12
plots$p13
plots$p23
```

```{r phase3-3d, eval=TRUE}
# Interactive 3D plot
rgl::open3d()
cols <- as.integer(factor(z))
rgl::plot3d(X, col = cols, size = 5, type = "s")
rgl::title3d("Simulated mixture on the unit sphere")
rgl::rglwidget()
```

# Attempting to fix mixtures of vMF for multiple K, and select K

We will fit vMF mixtures, and compute BIC (grid over K). BIC is used for
model selection, and helps us determine how well a model explains the
data. This will be perfect for determining whether to use vMF or
Euclidian models.

```{r fit-grid}
K_grid <- 1:8
fits <- vector("list", length(K_grid))
names(fits) <- paste0("K", K_grid)

# Data must be in a matrix. 

X_unit <- normalize_rows(X)

for (K in K_grid) {
  fits[[paste0("K", K)]] <- movMF::movMF(X_unit, k = K, nruns = 10)
}

# Extract, store BIC values
bic_df <- data.frame(
  K = K_grid,
  BIC = sapply(fits, BIC)
)

bic_df
```

```{r bic-plot}
ggplot(bic_df, aes(x = K, y = BIC)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(title = "Model selection for vMF mixtures", subtitle = "Choose K with the smallest BIC", x = "Number of components (K)", y = "BIC")
```

```{r choose-K}
K_hat <- bic_df$K[which.min(bic_df$BIC)]
K_hat
best_fit <- fits[[paste0("K", K_hat)]]
```

(i) BIC is determining what amount of models the data most likely has,
    with a lower BIC being better. In this case, BIC shows that 5 is the
    best fit.

(ii) This chosen K makes sense, as we did indeed select 5 mixtures to
     work with earlier. However, the BIC was nearly flat, which indicate
     that movRF was starting to having some trouble.

# Cluster Assignments, Visualization (vMF mixture)

```{r vmf-clusters}
vmf_cluster <- predict(best_fit)
vmf_cluster <- factor(vmf_cluster)
table(vmf_cluster)
```

```{r vmf-projections}
plots_vmf <- plot_projections(X_unit, label = vmf_cluster, title_prefix = paste0("vMF mixture clustering (K = ", K_hat, ")"))
plots_vmf$p12
plots_vmf$p13
plots_vmf$p23
```

```{r vmf-centers}
theta_hat <- best_fit$theta

kappa_hat <- apply(theta_hat, 2, function(x) norm(x,"2"))
kappa_hat

mu_hat <- apply(theta_hat, 2, function(x) x/norm(x,"2"))
mu_hat
```

movMF calculated five clusters here, all somewhat reminiscent of the
actual clusters we created in the simulation.

Simulated vMF cluster means: 1: Weight: 0.30, mean: (1, 0, 0), K: 25 2:
Weight: 0.25, mean: (0, 1, 0), K: 40 3: Weight: 0.20, mean: (0, 0, 1),
K: 30 4: Weight: 0.15, mean: (1, 1, 0), K: 15 5: Weight: 0.10, mean: (1,
0, 1), K: 20

Estimated vMF clusters means:

1: mean: (0.454, -0.005, 0.388) -\> Sim 5 2: mean: (0.000, 0.971, 0.005)
-\> Sim 2 3: mean: (0.336, 0.239, -0.0133) -\> Sim 4 4: mean: (0.023,
-0.010, 0.922) -\> Sim 3 5: mean: (0.824, 0.004, 0.004) -\> Sim 1

It is fairly obvious still which simulated cluster each estimated
cluster is mapping to. However, movMF is beginning to struggle a bit
with simulated mixtures 5 and 4, who have two directions they tend
towards.

------------------------------------------------------------------------

# Comparisons with Euclidean k-means

We will now test Euclidian k-means on the same data.

```{r kmeans-euclid}
set.seed(123)
km <- kmeans(X_unit, centers = K_hat, nstart = 25)
km_cluster <- factor(km$cluster)

table(km_cluster)
```

```{r km-projections}
plots_km <- plot_projections(X_unit, label = km_cluster, title_prefix = paste0("Euclidean k-means (K = ", K_hat, ")"))
plots_km$p12
plots_km$p13
plots_km$p23
```

### Agreement Metric

```{r agreement}
# Simple agreement table (labels might be permuted, so interpret carefully)
table(vmf_cluster, km_cluster)
```

An interpretation of this is what cluster corresponds to each cluster
from km to vmf. For example, km cluster 2 and vmf cluster 1 mostly
correspond to each other.

They appear to split the data slightly differently, but are mostly in
agreement. vMF 4 and km 3 are in total agreement, but there is some
variance with vMF 5.

# Comparative Study

In this study, we determine when spherical modeling outperforms
Euclidian clustering, and vice versa. To do this, a simulation grid with
varying sample sizes, concentration, and separation among cluster
centers is created. vMF and k-means models are ran against the
simulation grid.

```{r sim-grid}
library(dplyr)
handlers("txtprogressbar")  # simple, clean bar

simulate_mixture <- function(n, K, kappas, mus, w = rep(1/K, K)) {
  z <- sample(1:K, size = n, replace = TRUE, prob = w)
  X <- matrix(NA, nrow = n, ncol = length(mus[[1]]))
  for (k in 1:K) {
    nk <- sum(z == k)
    if (nk > 0) X[z == k, ] <- movMF::rmovMF(nk, theta = mus[[k]]*kappas[k])
  }
  list(X = normalize_rows(X), z = z)
}

within_cosine <- function(X, cluster) {
  cluster <- as.integer(factor(cluster))
  K <- max(cluster)
  out <- numeric(K)
  for (k in 1:K) {
    idx <- which(cluster == k)
    if (length(idx) <= 1) { out[k] <- NA; next }
    mu_hat <- colMeans(X[idx, , drop = FALSE])
    mu_hat <- mu_hat / sqrt(sum(mu_hat^2))
    out[k] <- mean(X[idx, , drop = FALSE] %*% mu_hat)
  }
  mean(out, na.rm = TRUE)
}

grid <- expand.grid(
  n = seq(100, 1000, by = 100),
  kappa = c(5, 10, 20, 30, 40, 50, 60, 70, 80, 90),
  rep = 1:10
)

grid <- as_tibble(grid)
grid
```

```{r sim-run}
plan(multisession, workers = 4)
library(progress)
pb <- progress_bar$new(
  total = nrow(grid),
  format = "  [:bar] :percent | ETA: :eta"
)

K <- 3
mus <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1)
)
mus <- lapply(mus, function(m) m / sqrt(sum(m^2)))

with_progress({
  p <- progressor(steps = nrow(grid))

  results <- future_lapply(
    seq_len(nrow(grid)),
    function(i) {
      p(sprintf("Sim %d", i))  # update progress bar

      row <- grid[i, ]

      sim <- simulate_mixture(
        n = row$n,
        K = K,
        kappas = rep(row$kappa, K),
        mus = mus
      )

      vmf_fit <- tryCatch(
        movMF::movMF(sim$X, k = K, nruns = 5),
        error = function(e) NULL
      )

      vmf_cluster <- if (is.null(vmf_fit)) NA else predict(vmf_fit)

      vmf_score <- if (all(is.na(vmf_cluster))) NA else
        within_cosine(sim$X, vmf_cluster)

      km <- kmeans(sim$X, centers = K, nstart = 10)
      km_score <- within_cosine(sim$X, km$cluster)

      data.frame(
        n = row$n,
        kappa = row$kappa,
        rep = row$rep,
        vmf_score = vmf_score,
        km_score = km_score
      )
    },
    future.seed = TRUE
  )

  results <- bind_rows(results)
})
```

```{r sim-summary}
summary_df <- results %>%
  group_by(n, kappa) %>%
  summarise(
    vmf_mean = mean(vmf_score, na.rm = TRUE),
    km_mean  = mean(km_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c(vmf_mean, km_mean),
                      names_to = "method", values_to = "score") %>%
  mutate(method = dplyr::recode(method, vmf_mean = "vMF mixture", km_mean = "Euclidean k-means"))

summary_df
```

```{r sim-plot}
ggplot(summary_df, aes(x = kappa, y = score, group = method, linetype = method, color = method)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ n, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Comparative study: spherical vs Euclidean clustering",
    subtitle = "Score = average within-cluster cosine similarity (higher is better)",
    x = "Concentration (kappa)",
    y = "Clustering score",
    color = "Method",
    linetype = "Method"
  )
```

Performance is nearly identical, it is very hard to tell the difference
between Euclidian vs vMF analysis. A deeper look at the accuracy has to
be taken, and across multiple mixture sizes. For now, however, let's just try to see the difference between the two lines.

```{r difference dataframe}
library(tidyr)

diff_df <- summary_df %>%
  pivot_wider(names_from = method, values_from = score) %>%
  mutate(
    diff = `vMF mixture` - `Euclidean k-means`,
    winner = ifelse(diff > 0, "vMF better", "k-means better")
  )
```

```{r plot ribbon}
ggplot(diff_df, aes(x = kappa)) +

  geom_ribbon(
    aes(
      ymin = ifelse(winner == "vMF better", `Euclidean k-means`, `vMF mixture`),
      ymax = ifelse(winner == "vMF better", `vMF mixture`, `Euclidean k-means`),
      fill = winner
    ),
    alpha = 0.6
  ) +

  geom_line(aes(y = `vMF mixture`, color = "vMF mixture"), linewidth = 0) +
  geom_line(aes(y = `Euclidean k-means`, color = "Euclidean k-means"), linewidth = 0) +

  geom_point(aes(y = `vMF mixture`, color = "vMF mixture"), size = 0.8) +
  geom_point(aes(y = `Euclidean k-means`, color = "Euclidean k-means"), size = 0.8) +

  facet_wrap(~ n, scales = "free_y") +
  theme_bw() +

  scale_fill_manual(values = c(
    "vMF better" = "violet",
    "k-means better" = "red"
  )) +

  labs(
    title = "Comparative study: spherical vs Euclidean clustering",
    subtitle = "Shaded area shows which method performs better",
    x = "Concentration (kappa)",
    y = "Clustering score",
    fill = "Winner",
    color = "Method"
  )
```
```{r win rate map}
win_df <- results %>%
  mutate(
    vmf_win = vmf_score > km_score
  ) %>%
  group_by(n, kappa) %>%
  summarise(
    win_rate = mean(vmf_win, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(win_df, aes(x = factor(kappa), y = factor(n), fill = win_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0.5,
    name = "vMF win rate"
  ) +
  labs(
    title = "Win Rate: vMF vs Euclidean k-means",
    subtitle = "Red = vMF dominates, Blue = k-means dominates",
    x = "Concentration (kappa)",
    y = "Sample size (n)"
  ) +
  theme_bw()
```
As we can see here, vMF seems to win across the board, with an exception with n=600 and a very low k value. A much deeper analysis needs to be done, so we will proceed with a win heatmap for various mixture sizes, k values, and sample sizes. This deep analysis will provide us with helpful usecases for when we need to use Euclidean vs movRF analysis.

