---
title: "R Notebook"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this file, a large scale comparison between Euclidian and sphericial analysis models will be conducted. 

- **Phase 5a: Mixture Modeling** The efficacy of Euclidian and spherical analysis models in   terms of fitting clusters will be conducted. 

- **Phase 5b: Comparative Study, mixed parameters** The efficacy of Euclidian and spherical analysis models will be compared against varied parameters; namely number of mixtures, sample size, and concentration. This section aims to give a final and in-depth look at when which model should be used.
Scenarios included are:
- Same weight: Each mixture uniformly has the same weight
- Varying weight: Mixtures have varying concentrations.
- Concentrations: Varying concentrations with each mixture
- Sample size: Varying sample sizes
- Randomized: Random parameters for each of the above, and analysis provided.
For each of these scenarios, a win rate heatmap is used to analyze. 2D graphs rarely have the minute detail required to tell which model is actually winning out.

```{r packages}
install.packages("rlang")
pkgs <- c("movMF", "Directional", "ggplot2", "dplyr", "tidyr", "rgl")
library(future.apply)
library(progressr)

plan(multisession, workers = 4)

handlers("txtprogressbar")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, require, character.only = TRUE))

has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (has_rgl) library(rgl)
```

```{r seed}
set.seed(123)
```

# Useful helper functions

```{r helpers}
# Normalize rows of a matrix to unit length
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norms <- sqrt(rowSums(X^2))
  return(X/norms)
}

# Check unit length (returns summary)
check_unit_length <- function(X) {
  r <- sqrt(rowSums(as.matrix(X)^2))
  return(summary(r))
}

# Angular distance between two unit vectors (in radians and degrees)
angle_between <- function(u, v) {
  u <- as.numeric(u) / sqrt(sum(u^2))
  v <- as.numeric(v) / sqrt(sum(v^2))
  val <- acos(max(min(sum(u*v), 1), -1))
  return(c(radians = val, degrees = val * 180 / pi))
}

# Fast 2D projections for 3D spherical data
plot_projections <- function(X, label = NULL, title_prefix = "Projection") {
  X <- as.data.frame(X)
  colnames(X) <- paste0("x", 1:ncol(X))
  if (!is.null(label)) X$cluster <- factor(label)

  p1 <- ggplot(X, aes(x = x1, y = x2, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x2")) +
    theme_bw()

  p2 <- ggplot(X, aes(x = x1, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x3")) +
    theme_bw()

  p3 <- ggplot(X, aes(x = x2, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x2 vs x3")) +
    theme_bw()

  return(list(p12 = p1, p13 = p2, p23 = p3))
}

make_kappas <- function(K, kappa_mean, variation = 1) {
  
  # map variation level to log-scale spread
  sigma <- switch(
    as.character(variation),
    "1" = 0.2,
    "2" = 0.5,
    "3" = 0.8,
    "4" = 1.2,
    stop("variation must be 1, 2, 3, or 4")
  )
  
  # log-mean so that expected value ≈ kappa_mean
  mu <- log(kappa_mean) - (sigma^2 / 2)
  
  # sample kappas
  kappas <- rlnorm(K, meanlog = mu, sdlog = sigma)
  
  return(kappas)
}

make_weights <- function(K, variation = 1) {
  
  alpha <- switch(
    as.character(variation),
    "1" = 10,
    "2" = 2,
    "3" = 0.7,
    "4" = 0.3,
    stop("variation must be 1, 2, 3, or 4")
  )
  
  # Dirichlet via Gamma
  x <- rgamma(K, shape = alpha, rate = 1)
  w <- x / sum(x)
  
  return(w)
}

generate_mus <- function(K, d = 3, separation = 1, max_iter = 500) {
  
  target <- switch(
    as.character(separation),
    "1" = c(0.2, 0.4),
    "2" = c(0.5, 0.7),
    "3" = c(0.75, 0.9)
  )
  
  for (iter in 1:max_iter) {
    M <- matrix(rnorm(K * d), K, d)
    M <- M / sqrt(rowSums(M^2))
    
    cos_vals <- (M %*% t(M))[upper.tri(M %*% t(M))]
    
    if (mean(cos_vals) >= target[1] && mean(cos_vals) <= target[2]) {
      return(split(M, row(M)))
    }
  }
  
  split(M, row(M))
}

```

------------------------------------------------------------------------

# Phase 5a: Varied Mixtures

We create a set of mixtures we can test against. These will also be phased in phase 5b.


```{r simulated mixture #2}
# Simulation
simulate_mixture <- function(n, K, kappas, mus, w) {
  z <- sample(1:K, size = n, replace = TRUE, prob = w)
  X <- matrix(NA, nrow = n, ncol = length(mus[[1]]))
  
  for (k in 1:K) {
    nk <- sum(z == k)
    if (nk > 0) {
      X[z == k, ] <- movMF::rmovMF(nk, theta = mus[[k]] * kappas[k])
    }
  }
  
  list(X = normalize_rows(X), z = z)
}

# Scoring
within_cosine <- function(X, cluster) {
  cluster <- as.integer(factor(cluster))
  K <- max(cluster)
  out <- numeric(K)
  
  for (k in 1:K) {
    idx <- which(cluster == k)
    if (length(idx) <= 1) next
    
    mu_hat <- colMeans(X[idx, , drop = FALSE])
    mu_hat <- mu_hat / sqrt(sum(mu_hat^2))
    out[k] <- mean(X[idx, , drop = FALSE] %*% mu_hat)
  }
  
  mean(out, na.rm = TRUE)
}

# =========================
# PARAMETER GRID
# =========================
grid <- expand.grid(
  n = seq(100, 1000, by = 200),
  kappa_mean = seq(10, 100, by = 20),
  weight_var = 1:4,
  kappa_var = 1:4,
  mu_sep = 1:3,
  rep = 1:50
) %>%
  as_tibble()

K <- 5
d <- 3

# =========================
# RUN SIMULATION
# =========================
with_progress({
  p <- progressor(steps = nrow(grid))
  
  results <- future_lapply(seq_len(nrow(grid)), function(i) {
    
    p(sprintf("Sim %d", i))
    row <- grid[i, ]
    
    w <- make_weights(K, row$weight_var)
    kappas <- make_kappas(K, row$kappa_mean, row$kappa_var)
    mus <- generate_mus(K, d, row$mu_sep)
    
    sim <- simulate_mixture(row$n, K, kappas, mus, w)
    
    vmf_fit <- tryCatch(
      movMF::movMF(sim$X, k = K, nruns = 3),
      error = function(e) NULL
    )
    
    vmf_cluster <- if (is.null(vmf_fit)) NA else predict(vmf_fit)
    
    vmf_score <- if (all(is.na(vmf_cluster))) NA else
      within_cosine(sim$X, vmf_cluster)
    
    km <- kmeans(sim$X, centers = K, nstart = 5)
    km_score <- within_cosine(sim$X, km$cluster)
    
    data.frame(
      n = row$n,
      kappa_mean = row$kappa_mean,
      weight_var = row$weight_var,
      kappa_var = row$kappa_var,
      mu_sep = row$mu_sep,
      rep = row$rep,
      vmf_score = vmf_score,
      km_score = km_score
    )
    
  }, future.seed = TRUE)
  
  results <- bind_rows(results)
})

# =========================
# SUMMARY
# =========================
results <- results %>%
  mutate(
    diff = vmf_score - km_score,
    vmf_win = vmf_score > km_score
  )

win_df <- results %>%
  group_by(n, kappa_mean, weight_var, kappa_var, mu_sep) %>%
  summarise(
    win_rate = mean(vmf_win, na.rm = TRUE),
    diff_mean = mean(diff, na.rm = TRUE),
    .groups = "drop"
  )

# =========================
# PLOT FOR EACH μ SEPARATION
# =========================
for (lvl in 1:3) {
  
  plot_df <- win_df %>%
    filter(mu_sep == lvl)
  
  p <- ggplot(plot_df, aes(x = factor(kappa_mean), y = factor(n), fill = win_rate)) +
    geom_tile(color = "white") +
    
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0.5,
      name = "vMF win rate"
    ) +
    
    facet_grid(weight_var ~ kappa_var) +
    
    labs(
      title = "vMF vs k-means Performance",
      subtitle = paste("μ separation level =", lvl,
                       "| Rows: weight variation | Columns: kappa variation"),
      x = "Mean concentration (kappa)",
      y = "Sample size (n)"
    ) +
    
    theme_bw()
  
  print(p)
}
```

```{r simulated mixture #4}
```

```{r simulated mixture #5}
```


