---
title: "Phase 3 & 4: vMF Mixtures, Clustering, and Comparative Study"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This R Markdown file provides **instructions + starter code** for:

- **Phase 3: Mixture Modeling and Clustering**  
  Fit mixtures of von Mises–Fisher (vMF) distributions, select the number of clusters (e.g., BIC), visualize clusters on the sphere, and compare with Euclidean k-means.

- **Phase 4: Comparative Study**  
  Run a simulation comparing Euclidean vs spherical models.

> Tip: Copy/paste results, plots, and interpretations into your final report sections:  
**(5) Mixture Modeling and Clustering** and **(6) Application or Simulation Study**.

---

# Required packages

```{r packages}
pkgs <- c("movMF", "Directional", "ggplot2", "dplyr", "tidyr", "rgl")

to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, require, character.only = TRUE))

has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (has_rgl) library(rgl)
```

```{r seed}
set.seed(123)
```

---

# Helper functions (useful in both phases)

```{r helpers}
# Normalize rows of a matrix to unit length
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norms <- sqrt(rowSums(X^2))
  return(X/norms)
}

# Check unit length (returns summary)
check_unit_length <- function(X) {
  r <- sqrt(rowSums(as.matrix(X)^2))
  return(summary(r))
}

# Angular distance between two unit vectors (in radians and degrees)
angle_between <- function(u, v) {
  u <- as.numeric(u) / sqrt(sum(u^2))
  v <- as.numeric(v) / sqrt(sum(v^2))
  val <- acos(max(min(sum(u*v), 1), -1))
  return(c(radians = val, degrees = val * 180 / pi))
}

# Fast 2D projections for 3D spherical data
plot_projections <- function(X, label = NULL, title_prefix = "Projection") {
  X <- as.data.frame(X)
  colnames(X) <- paste0("x", 1:ncol(X))
  if (!is.null(label)) X$cluster <- factor(label)

  p1 <- ggplot(X, aes(x = x1, y = x2, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x2")) +
    theme_bw()

  p2 <- ggplot(X, aes(x = x1, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x3")) +
    theme_bw()

  p3 <- ggplot(X, aes(x = x2, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x2 vs x3")) +
    theme_bw()

  return(list(p12 = p1, p13 = p2, p23 = p3))
}
```

---

# Phase 3: Mixture Modeling and Clustering

## Phase 3 deliverables (checklist)

By the end of Phase 3, you should have:

- A **mixture vMF model fit** (for one or multiple K values)
- A **model selection** plot/table (e.g., BIC vs K)
- **Cluster visualizations** (2D projections + 3D)
- A **comparison to Euclidean k-means** (same K), with a short interpretation

## Data for Phase 3

You have two options:

1. **Use simulated mixture data** (recommended starter).
2. **Use your Phase 2 simulated dataset** and *pretend* it is multi-group by mixing runs.

Below we generate an example simulated **mixture of vMF** in 3D.

### Step 1 — Simulate a mixture of vMF distributions

```{r simulate-mixture}
d <- 3
K_true <- 3
n <- 600

# Mixture weights (must sum to 1)
w <- c(0.35, 0.35, 0.30)

# Component mean directions (unit vectors)
mu_list <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1)
)
mu_list <- lapply(mu_list, function(m) m / sqrt(sum(m^2)))

# Component concentrations (kappa): larger => tighter clusters
kappa_list <- c(25, 15, 10)

# Sample component labels
z <- sample(1:K_true, size = n, replace = TRUE, prob = w)

# Generate data (rmovMF parameter: theta = mu*kappa)
X <- matrix(NA, nrow = n, ncol = d)
for (k in 1:K_true) {
  nk <- sum(z == k)
  if (nk > 0) {
    X[z == k, ] <- movMF::rmovMF(nk, theta = mu_list[[k]]*kappa_list[k]) 
  }
}

# Sanity checks
check_unit_length(X)
table(z)
```

### Step 2 — Visualize the mixture (2D projections + interactive 3D)

```{r phase3-projections}
plots <- plot_projections(X, label = z, title_prefix = "True mixture (simulated)")
plots$p12
plots$p13
plots$p23
```

```{r phase3-3d, eval=TRUE}
# Interactive 3D plot
rgl::open3d()
cols <- as.integer(factor(z))
rgl::plot3d(X, col = cols, size = 5, type = "s")
rgl::title3d("Simulated mixture on the unit sphere")
rgl::rglwidget()
```

---

## Fit mixtures of vMF (movMF) for multiple K and select K

### Step 3 — Fit vMF mixtures and compute BIC (grid over K)

`BIC`: https://en.wikipedia.org/wiki/Bayesian_information_criterion 

`movMF::movMF()` fits a mixture of vMF components using an EM-type algorithm.

```{r fit-grid}
K_grid <- 1:8
fits <- vector("list", length(K_grid))
names(fits) <- paste0("K", K_grid)

# Data must be a matrix; movMF expects unit vectors.
X_unit <- normalize_rows(X)

for (K in K_grid) {
  fits[[paste0("K", K)]] <- movMF::movMF(X_unit, k = K, nruns = 10)
}

# Extract and store BIC values
bic_df <- data.frame(
    K = K_grid,
    BIC = sapply(fits, BIC)
    )

bic_df
```

```{r bic-plot}
ggplot(bic_df, aes(x = K, y = BIC)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(title = "Model selection for vMF mixtures", subtitle = "Choose K with the smallest BIC", x = "Number of components (K)", y = "BIC")
```

```{r choose-K}
K_hat <- bic_df$K[which.min(bic_df$BIC)]
K_hat
best_fit <- fits[[paste0("K", K_hat)]]
```

> **Write-up prompt:**  
Explain (i) what BIC is doing here conceptually, and (ii) why your chosen K makes sense (also comment if BIC is flat/ambiguous).

---

## Cluster assignments and visualization (vMF mixture)

### Step 4 — Get cluster labels from the best vMF mixture

```{r vmf-clusters}
# Cluster assignments
# In movMF, each observation is assigned to the cluster with the maximum posterior probability
vmf_cluster <- predict(best_fit)
vmf_cluster <- factor(vmf_cluster)
table(vmf_cluster)
```

```{r vmf-projections}
plots_vmf <- plot_projections(X_unit, label = vmf_cluster, title_prefix = paste0("vMF mixture clustering (K = ", K_hat, ")"))
plots_vmf$p12
plots_vmf$p13
plots_vmf$p23
```

```{r vmf-centers}
# Extract estimated mean directions (centers) and concentration parameters (variances) for individual clusters
theta_hat <- best_fit$theta

kappa_hat <- apply(theta_hat, 2, function(x) norm(x,"2"))
kappa_hat

mu_hat <- apply(theta_hat, 2, function(x) x/norm(x,"2"))
mu_hat
```

> **Write-up prompt:**  
Briefly describe each cluster (size + where it sits on the sphere).  
If you simulated the data, compare estimated vs true centers and comment on mismatch.

---

## Compare with Euclidean k-means

### Step 5 — Run Euclidean k-means (same K) on the raw coordinates

This is a useful comparison because Euclidean k-means ignores spherical geometry.

```{r kmeans-euclid}
set.seed(123)
km <- kmeans(X_unit, centers = K_hat, nstart = 25)
km_cluster <- factor(km$cluster)

table(km_cluster)
```

```{r km-projections}
plots_km <- plot_projections(X_unit, label = km_cluster, title_prefix = paste0("Euclidean k-means (K = ", K_hat, ")"))
plots_km$p12
plots_km$p13
plots_km$p23
```

### Optional: simple agreement metric

```{r agreement}
# Simple agreement table (labels might be permuted, so interpret carefully)
table(vmf_cluster, km_cluster)
```

> **Write-up prompt:**  
Compare the vMF mixture clusters vs k-means:  
- Do they split the data similarly or differently?  
- Where does spherical modeling appear to help (or not)?

---

# Phase 4: Comparative Study

Phase 4 asks you to produce a **simulation comparison study**: Euclidean vs spherical across conditions.  

---

## Goal: quantify when spherical modeling outperforms Euclidean clustering.

### Step 1 — Define a simulation grid

We vary:
- sample size `n`
- concentration `kappa` (cluster tightness)
- separation among cluster centers (optional)

Then compare clustering quality for vMF mixtures vs k-means.

```{r sim-grid}
library(dplyr)

simulate_mixture <- function(n, K, kappas, mus, w = rep(1/K, K)) {
  z <- sample(1:K, size = n, replace = TRUE, prob = w)
  X <- matrix(NA, nrow = n, ncol = length(mus[[1]]))
  for (k in 1:K) {
    nk <- sum(z == k)
    if (nk > 0) X[z == k, ] <- movMF::rmovMF(nk, theta = mus[[k]]*kappas[k])
  }
  list(X = normalize_rows(X), z = z)
}

# Simple scoring: average within-cluster cosine similarity (higher is better).
within_cosine <- function(X, cluster) {
  cluster <- as.integer(factor(cluster))
  K <- max(cluster)
  out <- numeric(K)
  for (k in 1:K) {
    idx <- which(cluster == k)
    if (length(idx) <= 1) { out[k] <- NA; next }
    mu_hat <- colMeans(X[idx, , drop = FALSE])
    mu_hat <- mu_hat / sqrt(sum(mu_hat^2))
    out[k] <- mean(X[idx, , drop = FALSE] %*% mu_hat)
  }
  mean(out, na.rm = TRUE)
}

grid <- expand.grid(
  n = c(100, 300, 600),
  kappa = c(5, 10, 20, 30),
  rep = 1:10
)

grid <- as_tibble(grid)
grid
```

### Step 2 — Run the experiment

```{r sim-run}
K <- 3
mus <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1)
)
mus <- lapply(mus, function(m) m / sqrt(sum(m^2)))

results <- grid %>%
  rowwise() %>%
  mutate(
    sim = list(simulate_mixture(n = n, K = K, kappas = rep(kappa, K), mus = mus)),
    # vMF mixture clustering
    vmf_fit = list(movMF::movMF(sim$X, k = K,nruns = 10)),
    vmf_cluster = list(predict(vmf_fit)),
    vmf_score = within_cosine(sim$X, vmf_cluster),
    # Euclidean kmeans clustering
    km = list(kmeans(sim$X, centers = K, nstart = 10)),
    km_score = within_cosine(sim$X, km$cluster)
  ) %>%
  ungroup() %>%
  select(n, kappa, rep, vmf_score, km_score)

head(results)
```

### Step 3 — Summarize and visualize

```{r sim-summary}
summary_df <- results %>%
  group_by(n, kappa) %>%
  summarise(
    vmf_mean = mean(vmf_score, na.rm = TRUE),
    km_mean  = mean(km_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c(vmf_mean, km_mean),
                      names_to = "method", values_to = "score") %>%
  mutate(method = dplyr::recode(method, vmf_mean = "vMF mixture", km_mean = "Euclidean k-means"))

summary_df
```

```{r sim-plot}
ggplot(summary_df, aes(x = kappa, y = score, group = method, linetype = method)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ n, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Comparative study: spherical vs Euclidean clustering",
    subtitle = "Score = average within-cluster cosine similarity (higher is better)",
    x = "Concentration (kappa)",
    y = "Clustering score"
  )
```

> **Write-up prompt:**  
Summarize how performance changes with `n` and `kappa`.  
Give 2–3 concrete takeaways (e.g., “When clusters are diffuse (low kappa), vMF mixtures…").

---

## Phase 3 submission
- Your Phase 3 Rmd (HTML) with:
  - BIC vs K (or other selection rationale)
  - chosen K and clustering plots
  - k-means comparison and short interpretation

## Phase 4 submission
- Your Phase 4 Rmd (HTML) with:
  - Track A/B clearly stated
  - methods + results + discussion (~1–2 pages of writing)
  - at least **two limitations** and **one future direction**

---

# References 

- Hornik, K. and Grün, B. (2014). *movMF: An R package for fitting mixtures of von Mises–Fisher distributions.*
- Banerjee, A., Dhillon, I. S., Ghosh, J., and Sra, S. (2005). *Clustering on the Unit Hypersphere using vMF distributions.*
- Mardia, K. V. and Jupp, P. E. (2000). *Directional Statistics.*
- Sra, S. (2012). *A short note on parameter approximation for von Mises–Fisher distributions.*
