---
title: "R Notebook"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this file, a large scale comparison between Euclidian and sphericial analysis models will be conducted. 

- **Phase 5a: Mixture Modeling** The efficacy of Euclidian and spherical analysis models in   terms of fitting clusters will be conducted. 

- **Phase 5b: Comparative Study, mixed parameters** The efficacy of Euclidian and spherical analysis models will be compared against varied parameters; namely number of mixtures, sample size, and concentration. This section aims to give a final and in-depth look at when which model should be used.
Scenarios included are:
- Same weight: Each mixture uniformly has the same weight
- Varying weight: Mixtures have varying concentrations.
- Concentrations: Varying concentrations with each mixture
- Sample size: Varying sample sizes
- Randomized: Random parameters for each of the above, and analysis provided.
For each of these scenarios, a win rate heatmap is used to analyze. 2D graphs rarely have the minute detail required to tell which model is actually winning out.

```{r packages}
install.packages("rlang")
pkgs <- c("movMF", "Directional", "ggplot2", "dplyr", "tidyr", "rgl")

to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, require, character.only = TRUE))

has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (has_rgl) library(rgl)
```

```{r seed}
set.seed(123)
```

# Useful helper functions

```{r helpers}
# Normalize rows of a matrix to unit length
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norms <- sqrt(rowSums(X^2))
  return(X/norms)
}

# Check unit length (returns summary)
check_unit_length <- function(X) {
  r <- sqrt(rowSums(as.matrix(X)^2))
  return(summary(r))
}

# Angular distance between two unit vectors (in radians and degrees)
angle_between <- function(u, v) {
  u <- as.numeric(u) / sqrt(sum(u^2))
  v <- as.numeric(v) / sqrt(sum(v^2))
  val <- acos(max(min(sum(u*v), 1), -1))
  return(c(radians = val, degrees = val * 180 / pi))
}

# Fast 2D projections for 3D spherical data
plot_projections <- function(X, label = NULL, title_prefix = "Projection") {
  X <- as.data.frame(X)
  colnames(X) <- paste0("x", 1:ncol(X))
  if (!is.null(label)) X$cluster <- factor(label)

  p1 <- ggplot(X, aes(x = x1, y = x2, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x2")) +
    theme_bw()

  p2 <- ggplot(X, aes(x = x1, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x1 vs x3")) +
    theme_bw()

  p3 <- ggplot(X, aes(x = x2, y = x3, color = cluster)) +
    geom_point(alpha = 0.7) +
    labs(title = paste0(title_prefix, ": x2 vs x3")) +
    theme_bw()

  return(list(p12 = p1, p13 = p2, p23 = p3))
}
```

------------------------------------------------------------------------

# Phase 5a: Varied Mixtures

We create a set of mixtures we can test against. These will also be phased in phase 5b.

```{r simulated mixture #1}
d <- 3 # Number of dimensions
K_true <- 3 # Number of mixtures
# n is set to varying values

# Uniform mixture weights (for varying weights, custom values are set)
w <- c(0.33, 0.33, 0.34)

# Component mean directions
mu_list <- list(
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1),
)

# Component concentrations (kappas) can be made to varying difficulties
k_list_m1_d1 <- c(100, 110, 80)
k_list_m1_d1 <- c(50, 60, 40)
k_list_m1_d1 <- c(100, 110, 80)


```

```{r simulated mixture #2}
```

```{r simulated mixture #3}
```

```{r simulated mixture #4}
```

```{r simulated mixture #5}
```


